Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-05-29 parallel.tex
working on section 12.

Load balancing on levels other than the root level (i.e. grids for which the level $l > 0$) is different as the refined patches are not generally uniformly distributed.   Grids on refined levels are first placed on the same processor as their parent to minimize communication; however, this generally does not result in a well-balanced load.  Therefore, the code has a number of options for load balancing the grids on a given level. Each grid is assigned a load, generally equal to the number of cells in the grid.  The first load-balancing option is to move a grid from the processor with the highest load to the processor with the lowest load, with the proviso that only grids with load factors less than half the difference between the highest and lowest loaded processors will be moved.  This continues until the load ratio between the most-to-least loaded processor is below 1.05 or until no suitable grid can be found to transfer.  A second load balancing option uses a space-filling Hilbert curve to order the grids by their approximate spatial position.  Then, once the grids have a specific one-dimensional ordering, we can divide up the grids into $N_p$ groups (as equally as possible).  Load balancing is done separately for each level.  Clearly load balancing is most successful if there are significantly more grids than processors; however, small grids are less efficient (because of their ghost zones), and so the code uses a simple heuristic in order to split up grids until there are of order 10 grids per processor.  This generally results in good load-balancing while not producing grids which are wastefully small.

Communication between processors is done using a non-blocking communication strategy that allows overlap of communication and computation.  This can be done efficiently because each processor retains a copy of the entire hierarchy of grids, except that grids which do not `live' on a given processor only contain meta data (essentially location and size of the grid; such grids are denoted as `ghost' grids).   Replicating the hierarchy means that all communication of data from one grid to another can be identified by each processor independently.  The metadata for `ghost' grids are quite small and so the extra memory required is generally not onerous unless very large numbers of grids are used (more than a few hundred thousand grids).   A schematic of this distribution is shown in Figure~\ref{fig.amr_hierarchy}.

Data is transferred through a three-step procedure that takes advantage of MPI capabilities: (i) as the code progresses, and data is needed from another grid on another processor, the receiving processor posts an MPI non-blocking receive indicating that it is expecting data; this outstanding receive is recorded in a table , (ii) the sending processor calls the MPI non-blocking send function, and then finally (iii) the receiving processor, after it has carried out all the computation it can, waits for any MPI message to arrive.  Each message is coded so that it can be matched with appropriate receive posted in the first step, and based on that, the appropriate routine is called to processes the data.  Step (iii) is repeated until there are no outstanding receives.

@@

Because of the wide variety of simulations, methods, and uses of Enzo,
it is difficult to define exactly which routines are most costly
during a given simulation.  As such, we have designed a lightweight
registering system that has been implemented for the most commonly
used routines (such as the hydrodynamic and gravity solvers) as well
as refinement level timers that measure the (exclusive) time spent on
each level.  Beyond this minimal set of routines, we have designed a
simple way for the user to modify the source by adding
\texttt{TIMER\_START(``Your Routine Name'')} and
\texttt{TIMER\_END(``Your Routine Name'')}.  These timers are
automatically registered in a
std::map\footnote{http://www.cplusplus.com/reference/map/map/}.  These
timers are created and managed individually on each processor in an
asynchronous fashion.

At each complete EvolveHierarchy (or less often if specified), each timer is
then communicated to the root processor where it calculates the mean, standard
deviation, minimum, and maximum for each of the timers across all processors. 
For level timers, there are also attributes such as the number of cell updates,
the current number of grids, and the average cells/s/MPI process.  This
information is then output to a ``performance.out'' logfile.  This provides a 
simplified interface to the user that can be used to diagnose performance 
issues as well as estimate a given problem type's scalability.  In addition to 
the logfile, we have developed a plotting interface for quickly producing 
figures that process the data from the logfile.  These capabilities are 
described in the online documentation, along with a further discussion of the 
performance measurement implementation.

@@

Figure~\ref{fig.uniscale} shows \enzo\ weak scaling results for a
sequence of scaled unigrid Lyman alpha forest calculations. These
calculations include dark matter dynamics, hydrodynamics using the
piecewise parabolic method, six-species non-equilibrium chemistry and
radiative cooling, and a uniform metagalactic ultraviolet background.
In this sequence of test calculations, we perform a weak scaling test
on up to 13,824 MPI tasks on the NICS Kraken XT4 and ORNL Jaguar XT4
supercomputers\footnote{These simulations were performed in 2008,
prior to conversion of both machines to the current-generation
systems}.  In this test, each MPI task was given a $128^3$ root grid
tile (i.e., $128^3$ grid cells containing baryon quantities) and,
initially, approximately $128^3$ dark matter particles.  The number of
grid cells was constant throughout the calculation; the number of dark
matter particles varies as they are moved from subvolume to subvolume
as structure evolves.  The grid resolution was kept at a constant
comoving size of $\simeq 40$~kpc/h, and as the core count was
increased, so was the simulation volume.  On each machine, a compute
node contained a single AMD Opteron quad-core chip (2.1 Ghz on Jaguar;
2.3 Ghz on Kraken) with 2 GB/memory per core (8 GB/total per node).
Both machines used the SeaStar2 interconnect.  In the scaling study,
calculations were run with 1, 2 or 4 MPI tasks per node.  The figure
shows cell updates per second per MPI process; perfect scaling would
be a horizontal line.

As can be seen in Figure~\ref{fig.uniscale}, the unigrid weak scaling
performance of the code is extremely good for this problem, with only
a 20\% decrease in cell updates/second/task as the code is scaled from
8 to 4,096 MPI tasks, and a 40\% decrease in performance overall going
from 8 to 13,824 (or $24^3$) MPI tasks.  We speculate that this
decrease is likely to be partially due to global MPI communications
used to, e.g., calculate the overall timestep of the simulation, and
also likely due to load imbalances due to increasing cosmological
power (and thus an increasingly uneven distribution of dark matter
particles between MPI tasks at late times) as the simulation volume
grows.  We also observe that a systematic difference in speed can be
seen between the two machines, which can be attributed primarily to
the slightly faster CPUs on Kraken (2.3 Ghz, vs. 2.1 Ghz on Jaguar).
The difference in speed when using different numbers of MPI tasks per
node can be attributed primarily due to differences in competing usage
of shared cache on the quad-core chips used on this machine.

Broadly, excellent scaling in \enzo's unigrid mode is seen for a
variety of problems as long as each compute core is given an adequate
amount of work to do.  For cosmological simulations, this seems to be
roughly $128^3$ cells per core.  If fewer cells per core are used, the
CPU is essentially data-starved, and poor scaling is observed due to
computing units being idle while waiting for information to be
communicated from other processes (for, e.g., boundary information or
gravity solves).  Substantially larger cell counts per core would in
principle help scaling by reducing the amount of inter-process
communication needed, but larger cell counts are typically impractical
on most machines due to memory limits.

@@

Many astrophysical problems, such as cosmological galaxy formation
\citep{}, high-resolution disk galaxy simulations \citep{},
high-redshift \citep{} and present-day star formation \citep{},
involve multi-scale physics that span several orders of magnitude in
both space and time.  In these situations, using Enzo in its adaptive
mesh refinement mode is beneficial.  Because the adaptive grid
hierarchy is dynamic and not static, grid boundaries and, thus,
communication patterns can be unpredictable, hindering good strong
scaling at high core counts.

Figure \ref{fig:strong_scaling} shows strong scaling results from a
single 50 Mpc/h cosmology simulation run on $N_{\rm core}$ compute
cores, ranging from 128 to 16,384 at power-of-two intervals.  It was
run on the NICS Kraken XT5 supercomputer that has two AMD Opteron
hexa-core processors with 16 GB of memory per compute node.  The
simulations that utilized 128, 256, and 512 cores were executed on 128
nodes because of the memory requirements.  The higher core count jobs
were run with 8 cores per node.  This simulation would not run with 12
cores per node because of the memory overhead associated with the grid
hierarchy being duplicated on each MPI process.  However, this
overhead is greatly diminished if a hybrid-parallel (MPI + OpenMP)
approach is used.

This simulation includes dark matter dynamics, hydrodynamics using the
piecewise parabolic method, six-species non-equilibrium chemistry and
radiative cooling, and a uniform metagalactic ultraviolet background.
The simulation uses the space-filling curve method for load balancing
the AMR grids.  It has a $512^3$ root grid that is divided into 512
tiles, and 6 additional AMR levels are used.  We perform these scaling
tests when the simulation reaches $z=4$.  The results shown in Figure
\ref{fig:strong_scaling} come from a single root-level timestep
$\Delta t = 2.1\, \textrm{Myr}$.  At this time, there are $3.04 \times
10^5$ AMR grids, $9.03 \times 10^8$ ($\sim 967^3$) computational
cells, and $1.34 \times 10^8$ dark matter particles in total.  The
breakdown of the number of AMR grids, cells, timesteps, and number of
cell updates on each level is shown in Table \ref{tab:amr_scale}.

@@

The left panel in Figure \ref{fig:strong_scaling} shows the
computational and communication time spent on each level.  In the AMR
levels, there exists good strong scaling up to 2,048 cores, and
marginal speed-ups are found at 4,096 cores.  On the root-grid level,
there exists good scaling up to 1,024 cores, but the performance
dramatically decreases at higher core counts.  This occurs because the
root grid is not re-partitioned into $N_{\rm core}$ tiles when the
simulation is restarted with a different core count.  This feature can
be easily implemented and is planned in the next major release of
Enzo, where scaling results would be similar to the weak scaling shown
in \S\ref{sec:weak_scaling}.  The right panel in Figure
\ref{fig:strong_scaling} shows the time spent in some representative
major routines in Enzo.  The local physics routines, for example
\texttt{SolveHydroEquations}, exhibit perfect strong scaling because
they involve no communication.  By investigating the scaling behavior
in each routine, it is clear that the communication in the
\texttt{SetBoundaryConditions}, \texttt{SolveForPotential} (multi-grid
solver in AMR levels), and \texttt{RebuildHierarchy} are responsible
for the lack of strong scaling at $N_{\rm core} \ga$~4,096 in this
particular simulation.  The transpose of the root grid tiles are
responsible for the performance decrease because it is not optimized
for situations where the number of tiles is greater than the number of
MPI processes.  These results are directly applicable to simulations
with similar computational demands.  In simulations with less
computational cells, strong scaling will cease at a smaller core count
because the CPUs will become data-starved faster, and the opposite
occurs with larger simulations.

@@

In this section, we develop a simple, approximate model to estimate the compute time and memory required to complete a given calculation.  Note that we are {\it not} trying to model parallel performance (which we discussed in the previous section), but have an even simpler goal: to determine how to scale compute time estimates as we increase the resolution.  This is a straightforward thing to do in codes which use static grids, as the computational effort per time step is constant.  However, for an AMR calculation, the number of grids that will be generated during the run is not known in advance, and so the cpu time per problem time can vary drastically throughout the calculation.  
%For example, at the beginning, when the densities are nearly uniform, only the static grid is required and the calculation progresses rapidly.  However, as structure forms and dense clumps are generated, the number of grid points swells by orders of magnitude (an increase by $10^3$ is not uncommon) and most of the cpu time is consumed at late times.  Therefore, simply performing a few steps does produce a good estimate of the cpu time.

Nevertheless, we can try to determine how the compute time scales for a given run, as we increase the resolution.  First, we neglect particles and concentrate on the time taken by the grids, which can be justified both theoretically and empirically\footnote{Since the potential is calculated on the grid, the only particle costs are: depositing mass to the grid, interpolating accelerations, and updating particle positions and velocities.  These are relatively cheap operations.}.  Second, we break the problem down slightly, and examine the scaling over a short enough period of time that the grid structure does not change significantly (i.e. the number of grids at each level).  We then assume that the whole run scales in the same way, or in other words, that changes in the resolution affect the simulation in the same way at each time step.  This is usually a good approximation as the most costly part of the calculation typically don't change their characteristics substantially.

To make progress, we assume that the cost to advance a single cell by one timestep is a constant.  This can be incorrect if the chemistry solver requires many iterations, but is usually fairly accurate.  For a unigrid calculation, the time would by proportional to $N_{\rm root}^4$ since the number of cells scales as $N_{\rm root}^3$ and, assuming the Courant condition is the time limiting factor, then the number of steps to advance a given time is proportional to $N_{\rm root}.$  Therefore, to advance a hierarchy a given time, we find, accounting for all levels and using a refinement factor of 2.
\begin{equation}
t_{\rm SU} = C_1 \sum_l f_l N_{\rm root}^4 2^{4l}
\end{equation}
where $C_1$ is a constant, which can be thought of as the time taken to advance a single cell.  The factor $f_l$ is the fraction of the volume on a given level that is actually refined.  By definition $f_0 = 1$, and $f_l \le 1$.  In writing down this equation, we neglect a number of costs which are not directly proportional to cell count, including communication between processors, the $\log{N}$ factor for the root grid FFT, Cache misses, optimizations, and other costs associated with processing the hierarchy.  The first item on this list, in particular, is clearly not true for large processor counts; however, we neglect parallel considerations in this section.

Note that unless $f_l/f_{l+1} < 1/16$ (i.e. if less then about 6\% of a given level is further refined), the cost per level will increase with level.  A key question therefore, is the values of $f_l$.  Unfortunately, this depends strongly on the simulation being run.  In Figure~\ref{fig:scaling}, we show the values of $f_l$ for three simulations: two cosmological runs with varying box size, and a third simulation focusing on a single disk galaxy.

@@

As, the figure demonstrates, the $f_l$ values all show a sharp decline with level, dropping as power-laws with the level $l$.  Focusing first on the ``refine-everywhere" cosmological runs, which both show similar behavior despite the different box sizes and redshifts at which we collect the data.  In fact, we find the results are very typical for cosmological runs, with only slight variations depending on if we are simulating first stars or clusters of galaxies.  The top-left panel shows the ratio of $f_{l}/f_{l+1}$ which, remarkably, hovers around the critical value of 0.0625 determined earlier.  This implies that the relative cpu usage of each level, shown in the lower-left panel, is nearly flat, and that going to high levels of refinement only increases the amount of calculation by a factor of roughly $1/l_{\rm max}$.  Even in the most extreme case, the L80 run, the last level adds only 25\% to the computation time.

@@

The physical reason for the $f_{l+1}/f_l$ ratios found in the cosmological simulations appears to be due to a combination of the density structure of individual clouds, and the distribution of the clouds themselves.  In particular, we note that for a $\rho \propto r^{-2}$ profile, the Lagrangian refinement criteria typically used in such simulations produces an $f_{l+1}/f_l$ ratio of 1/8 for a single, resolved cloud.  Of course, at some point we would resolve the flat density of individual clouds and the ratio would climb, making further refinement more costly; however, we are not yet in this regime in this example.

@@

Finally, we have so far focused on scaling when increasing $l_{\rm max}$, but without changing the refinement criteria.  However, we can also keep the levels fixed and modify the refinement criteria.  For cosmological runs, this is typically done by increasing the mass resolution (and simultaneously decreasing the dark matter particle mass), and adding additional linear small-scale modes to the initial conditions.  The effect of this is to boost the $f_l$ values at all levels (except for the root grid where $f_0$ is already 1) by a factor linearly proportional to the increase in the mass resolution.  This is because the refinement criteria typically boosts the number of refined cells on the first and subsequent levels by this factor.  Again, we have empirically demonstrated that this scaling is approximately valid (provided we keep the maximum level constant).

Putting this together, we obtain an approximate scaling for cosmological simulations as 
\begin{equation}
t_{\rm SU} \propto M_{\rm res}^{-1} l_{\rm max}.
\end{equation}
which at least gives users some way to estimate compute times.

>>>>

Load balancing on levels other than the root level (i.e., grids for
which the level $l > 0$) is different as the refined patches are not
generally uniformly distributed.  Grids on refined levels are first
placed on the same processor as their parent to minimize
communication; however, this generally does not result in a
well-balanced load.  Therefore, the code has a number of options for
load balancing the grids on a given level. Each grid is assigned a
load, generally equal to the number of cells in the grid (which,
empirically, is a good estimate of computational cost).  The first
load-balancing option is to move a grid from the processor with the
highest load to the processor with the lowest load, with the proviso
that only grids with load factors less than half the difference
between the highest and lowest loaded processors will be moved.  This
continues until the load ratio between the most-to-least loaded
processor is below 1.05 or until no suitable grid can be found to
transfer.  A second load balancing option uses a space-filling Hilbert
curve to order the grids by their approximate spatial position.  Then,
once the grids have a specific one-dimensional ordering, we can divide
up the grids into $N_p$ groups (with the division taking place as
equally as possible).  Load balancing is done separately for each
level.  Clearly load balancing is most successful if there are
significantly more grids than processors; however, small grids are
less efficient (because of their ghost zones), and so the code uses a
simple heuristic in order to split up grids until there are of order
10 grids per processor.  This generally results in good load-balancing
while not producing grids that are wastefully small.

Communication between processors is done using a non-blocking
communication strategy that allows overlap of communication and
computation.  This can be done efficiently because each processor
retains a copy of the entire hierarchy of grids, except that grids
that do not `live' on a given processor only contain meta data
(essentially location and size of the grid; such grids are denoted as
`ghost' grids).  Replicating the hierarchy means that all
communication of data from one grid to another can be identified by
each processor independently.  The metadata for `ghost' grids are
quite small and so the extra memory required is generally not onerous
unless very large numbers of grids are used (more than a few hundred
thousand grids).  A schematic of this distribution is shown in
Figure~\ref{fig.amr_hierarchy}.

Data is transferred through a three-step procedure that takes
advantage of the capabilities of the MPI library: (i) as the code
progresses, and data is needed from another grid on another processor,
the receiving processor posts an MPI non-blocking receive indicating
that it is expecting data; this outstanding receive is recorded in a
table, (ii) the sending processor calls the MPI non-blocking send
function, and then finally (iii) the receiving processor, after it has
carried out all the computation it can, waits for any MPI message to
arrive.  Each message is coded so that it can be matched with the
appropriate receive posted in the first step, and based on that, the
appropriate routine is called to processes the data.  Step (iii) is
repeated until there are no outstanding receives.

@@

Because of the wide variety of simulations, methods, and uses of Enzo,
it is difficult to state generally which routines will be most costly
during a given simulation.  As such, we have designed a lightweight
registering system that has been implemented for the most commonly
used routines (such as the hydrodynamic and gravity solvers) as well
as refinement level timers that measure the time spent on each level.
Beyond this minimal set of routines, we have designed a simple way for
the user to modify the source by adding \texttt{TIMER\_START(``Your
Routine Name'')} and \texttt{TIMER\_END(``Your Routine Name'')}.
These timers are created and managed individually on each processor in
an asynchronous fashion, and contribute minimal computational, memory,
and IO overhead.

At each complete root grid time step (or less often if specified),
each timer is then communicated to the root processor where it
calculates the mean, standard deviation, minimum, and maximum for each
of the timers across all processors.  For level timers, there are also
attributes such as the number of cell updates, the current number of
grids, and the average cells/s/MPI process.  This information is then
output to a logfile.  This provides a simplified interface to the user
that can be used to diagnose performance issues as well as estimate a
given problem type's scalability.  In addition to the logfile, we have
developed a plotting interface for quickly producing figures that
process the data from the logfile.  These capabilities are described
in the online documentation, along with a further discussion of the
performance measurement implementation.

@@

Figure~\ref{fig.uniscale} shows \enzo\ weak scaling results for a
sequence of scaled unigrid Lyman alpha forest calculations. These
calculations include dark matter dynamics, hydrodynamics using the
piecewise parabolic method, six-species non-equilibrium chemistry and
radiative cooling, and a uniform metagalactic ultraviolet background.
In this sequence of test calculations, we perform a weak scaling test
on up to 13,824 MPI tasks on the NICS Kraken XT4 and ORNL Jaguar XT4
supercomputers\footnote{These simulations were performed prior to
conversion of both machines to the current-generation systems}.  In
this test, each MPI task was given a $128^3$ root grid tile (i.e.,
$128^3$ grid cells containing baryon quantities) and, initially,
approximately $128^3$ dark matter particles.  The number of grid cells
was constant throughout the calculation; the number of dark matter
particles varies as they are moved from subvolume to subvolume as
structure evolves.  The grid resolution was kept at a constant
comoving size of $\simeq 40$~kpc/h, and as the core count was
increased, so was the simulation volume.  On each machine, a compute
node contained a single AMD Opteron quad-core chip (2.1 Ghz on Jaguar;
2.3 Ghz on Kraken) with 2 GB/memory per core (8 GB/total per node).
Both machines used the SeaStar2 interconnect.  In the scaling study,
calculations were run with 1, 2, or 4 MPI tasks per node.  The figure
shows cell updates per second per MPI process; perfect scaling would
be a horizontal line.

As can be seen in Figure~\ref{fig.uniscale}, the unigrid weak scaling
performance of the code is extremely good for this problem, with only
a 20\% decrease in cell updates/second/task as the code is scaled from
8 to 4,096 MPI tasks, and a 40\% decrease in performance overall going
from 8 to 13,824 (or $24^3$) MPI tasks.  We speculate that this
decrease is likely to be partially due to global MPI communications
used to, e.g., calculate the overall timestep of the simulation, and
also likely due to load imbalances due to increasing cosmological
power (and thus an increasingly uneven distribution of dark matter
particles between MPI tasks at late times) as the simulation volume
grows.  We also observe that a systematic difference in speed can be
seen between the two machines, which can be attributed primarily to
the slightly faster CPUs on Kraken at th etime (2.3 Ghz, vs. 2.1 Ghz
on Jaguar).  The difference in speed when using different numbers of
MPI tasks per node can be attributed primarily due to differences in
competing usage of shared cache on the quad-core chips used on this
machine.

Broadly, excellent scaling in \enzo's unigrid mode is seen for a
variety of problems as long as each compute core is given an adequate
amount of work to do.  For cosmological simulations, this value has
been empirically determined to be roughly $128^3$ cells per core.  If
fewer cells per core are used, the CPU is essentially data-starved,
and poor scaling is observed due to computing units being idle while
waiting for information to be communicated from other processes (for,
e.g., boundary information or gravity solves).  Substantially larger
cell counts per core would in principle help scaling by reducing the
amount of inter-process communication needed, but larger cell counts
are typically impractical on most machines due to memory limits.

@@

Many astrophysical problems, such as cosmological galaxy formation
\citep{2012ApJ...749..140H}, high-resolution disk galaxy simulations
\citep{}, high-redshift \citep{2009Sci...325..601T}, and present-day
star formation \citep{Collins12a}, involve multi-scale physics that
span several orders of magnitude in both space and time.  In these
situations, using Enzo in its adaptive mesh refinement mode is
beneficial.  Because the adaptive grid hierarchy is dynamic, grid
boundaries and, thus, communication patterns can be unpredictable,
hindering strong scaling behavior to high core counts.

Figure \ref{fig:strong_scaling} shows strong scaling results from a
single 50 Mpc/h cosmology simulation run on $N_{\rm core}$ compute
cores, ranging from 128 to 16,384 cores at power-of-two intervals.  It
was run on the NICS Kraken XT5 supercomputer, which has two AMD
Opteron hexa-core processors with 16 GB of memory per compute node.
The simulations that utilized 128, 256, and 512 cores were executed on
128 nodes because of the memory requirements.  The higher core count
simulations were run with 8 cores per node.  This simulation would not
run with 12 cores per node because of the memory overhead associated
with the grid hierarchy being duplicated on each MPI process.
However, this overhead is greatly diminished if a hybrid-parallel (MPI
+ OpenMP) approach is used.

This simulation includes dark matter dynamics, hydrodynamics using the
piecewise parabolic method, six-species non-equilibrium chemistry and
radiative cooling, and a uniform metagalactic ultraviolet background.
The simulation uses the space-filling curve method for load balancing
the AMR grids.  It has a $512^3$ root grid that is divided into 512
tiles, and 6 additional AMR levels are used.  We perform these scaling
tests when the simulation reaches $z=4$, where the AMR grid hierarchy
is well-developed and is thus a reasonable representation of AMR
simulation behavior in general.  The results shown in Figure
\ref{fig:strong_scaling} come from a single root-level timestep
of $\Delta t = 2.1\, \textrm{Myr}$.  At this time, there are $3.04 \times
10^5$ AMR grids, $9.03 \times 10^8$ ($\sim 967^3$) computational
cells, and $1.34 \times 10^8$ dark matter particles in total.  The
breakdown of the number of AMR grids, cells, timesteps, and number of
cell updates on each level is shown in Table \ref{tab:amr_scale}.

@@

The left panel in Figure \ref{fig:strong_scaling} shows the
computational and communication time spent on each level.  In the AMR
levels, there exists good strong scaling up to 2,048 cores, and
marginal speed-ups are found at 4,096 cores.  On the root-grid level,
there exists good scaling up to 1,024 cores, but the performance
dramatically decreases at higher core counts.  This occurs because the
root grid is not re-partitioned into $N_{\rm core}$ tiles when the
simulation is restarted with a different core count.  This feature can
be easily implemented and is planned in the next major release of
Enzo, where scaling results would be similar to the weak scaling shown
in \S\ref{sec:weak_scaling}.  The right panel in Figure
\ref{fig:strong_scaling} shows the time spent in some representative
major routines in Enzo.  The local physics routines, for example
\texttt{SolveHydroEquations}, exhibit perfect strong scaling because
they involve no communication.  By investigating the scaling behavior
in each routine, it is clear that the communication in the
\texttt{SetBoundaryConditions}, \texttt{SolveForPotential} (multi-grid
solver in AMR levels), and \texttt{RebuildHierarchy} are responsible
for the lack of strong scaling at $N_{\rm core} \ga$~4,096 in this
particular simulation.  The transpose of the root grid tiles are
responsible for the performance decrease because it is not optimized
for situations where the number of tiles is greater than the number of
MPI processes.  These results are directly applicable to simulations
with similar computational demands.  In simulations with fewer
computational cells, strong scaling will cease at a smaller core count
because the CPUs will become data-starved more quickly, and the
opposite occurs with larger simulations.

@@

In this section, we develop a simple, approximate model to estimate
the compute time and memory required to complete a given calculation.
Note that we are {\it not} trying to model parallel performance (which
was discussed in the previous section), but have an even simpler goal:
to determine how to scale compute time estimates as we increase the simulation
resolution.  This is a straightforward thing to do in codes that use
static grids, as the computational effort per time step is constant.
However, for an AMR calculation, the number of grids that will be
generated during the run is not known in advance, and so the CPU time
per problem time can vary drastically throughout the calculation.
For example, at the beginning of the simulation when the densities are nearly uniform,
only the static grid is required and the calculation progresses
rapidly.  However, as structure forms and dense clumps are generated,
the number of grid points swells by orders of magnitude (an increase
of $10^3$ is not uncommon) and most of the cpu time is consumed at
late times.  Therefore, simply performing a few steps does produce a
good estimate of the cpu time.

Nevertheless, we can try to determine how the compute time scales for
a given run as we increase the resolution.  First, we neglect
particles and concentrate on the time taken by the grids, which can be
justified both theoretically and empirically\footnote{Since the
potential is calculated on the grid, the only particle costs are:
depositing mass to the grid, interpolating accelerations, and updating
particle positions and velocities.  These are relatively
computationally inexpensive operations.}.  Second, we break the
problem down slightly, and examine the scaling over a short enough
period of time that the grid structure does not change significantly
(i.e. the number of grids at each level remains approximately
constant).  We then assume that the whole run scales in the same way,
or in other words, that changes in the resolution affect the
simulation in the same way at each time step.  This is usually a good
approximation, as the most costly parts of the calculation typically
don't change their characteristics substantially between timesteps.

To make progress, we assume that the computational cost to advance a
single cell by one timestep is a constant.  This can be incorrect if
the chemistry solver requires many iterations, but is usually fairly
accurate.  For a unigrid calculation, the time would by proportional
to $N_{\rm root}^4$ since the number of cells scales as $N_{\rm
root}^3$ and, assuming the Courant condition is the time limiting
factor, then the number of steps to advance a given time is
proportional to $N_{\rm root}.$ Therefore, to advance a hierarchy a
given time, we find, accounting for all levels and using a refinement
factor of 2.
\begin{equation} t_{\rm SU} = C_1 \sum_l f_l N_{\rm root}^4 2^{4l}
\end{equation} where $C_1$ is a constant, which can be thought of as
the time taken to advance a single cell.  The factor $f_l$ is the
fraction of the volume on a given level that is actually refined.  By
definition $f_0 = 1$, and $f_l \le 1$.  In writing down this equation,
we neglect a number of costs which are not directly proportional to
cell count, including communication between processors, the $\log{N}$
factor for the root grid FFT, Cache misses, optimizations, and other
costs associated with processing the hierarchy.  The first item on
this list, in particular, is clearly not true for large processor
counts; however, we neglect parallel considerations in this section.

Note that unless $f_l/f_{l+1} < 1/16$ (i.e. if less then about 6\% of
a given level is further refined), the cost per level will increase
with level.  A key question, therefore, is the value of $f_l$ for each
level.  Unfortunately, this depends strongly on the simulation being
run.  In Figure~\ref{fig:scaling}, we show the values of $f_l$ for
three simulations: two cosmological runs with varying box size, and a
third simulation focusing on a single disk galaxy.

@@

As, the figure demonstrates, the $f_l$ values all show a sharp decline
with level, dropping as power-laws with the level $l$.  Focusing first
on the ``refine-everywhere" cosmological runs, which both show similar
behavior despite the different box sizes and redshifts at which we
collect the data.  In fact, we find the results are very typical for
cosmological runs, with only slight variations depending on if we are
simulating first stars or clusters of galaxies.  The top-left panel
shows the ratio of $f_{l}/f_{l+1}$ which, remarkably, hovers around
the critical value of 0.0625 determined earlier.  This implies that
the relative cpu usage of each level, shown in the lower-left panel,
is nearly flat, and that going to additional levels of refinement only
increases the amount of calculation by a factor of roughly $1/l_{\rm
max}$.  Even in the most extreme case, the L80 run, the last level
adds only 25\% to the computation time.

@@

The physical reason for the $f_{l+1}/f_l$ ratios found in the
cosmological simulations appears to be due to a combination of the
density structure of individual clouds, and the distribution of the
clouds themselves.  In particular, we note that for a $\rho \propto
r^{-2}$ density profile, the Lagrangian refinement criteria typically used in
such simulations produces an $f_{l+1}/f_l$ ratio of 1/8 for a single,
resolved cloud.  Of course, at some point we would resolve the flat
density of individual clouds and the ratio would climb, making further
refinement more costly; however, we are not yet in this regime in this
example.

@@

We have focused thus far on scaling when increasing $l_{\rm max}$, but
without changing the refinement criteria.  However, we can also keep
the levels fixed and modify the refinement criteria.  For cosmological
runs, this is typically done by increasing the mass resolution (and
simultaneously decreasing the dark matter particle mass), and adding
additional linear small-scale modes to the initial conditions.  The
effect of this is to boost the $f_l$ values at all levels (except for
the root grid where $f_0$ is already 1) by a factor linearly
proportional to the increase in the mass resolution.  This is because
the refinement criteria typically boosts the number of refined cells
on the first and subsequent levels by this factor.  Again, we have
empirically demonstrated that this scaling is approximately valid,
provided that we keep the maximum level constant.

Putting this together, we find an approximate scaling for the
computational time required for cosmological simulations:
\begin{equation}
t_{\rm SU} \propto M_{\rm res}^{-1} l_{\rm max}.
\end{equation}
This is approximately accurate, and gives users some way to estimate
compute times for cosmological calculations.  Similar scaling
relations can be obtained for other types of \enzo\ AMR simulations.
